---
title: "BivariateStats"
output: html_document
date: "2025-04-17"
---

```{r}
library(dplyr)
library(tidyr)
```

Descriptive stats:
```{r}
NHANES_cancer_project %>% 
  summarize(mean(AvgCigSmoked), sd(AvgCigSmoked), count=n())
NHANES_cancer_project %>% 
  drop_na(FamilyInc.Poverty) %>% 
  summarize(mean(FamilyInc.Poverty), sd(FamilyInc.Poverty), count=n())
```

factor: 
```{r}
library(questionr)

#count(database, variable)
#factor(C(database$variable, order=TRUE)

factor(c(NHANES_cancer_project$CancerType, NHANES_cancer_project))
factor(c(NHANES_cancer_project$CancerYN, NHANES_cancer_project))
factor(c(NHANES_cancer_project$AnnualHHInc, order=TRUE))
factor(NHANES_cancer_project$HowHealthy, order=TRUE, level=c("Excellent, VGood", "Good", "Fair", "Poor"))
```

Frequency tables:
```{r}
count(NHANES_cancer_project, CancerType)
count(NHANES_cancer_project, CancerYN)
count(NHANES_cancer_project, HowHealthy)
count(NHANES_cancer_project, AnnualHHInc)
```

```{r}
library(ggplot2)
```

Bar Charts for nominal and ordinal:
```{r}
#ggplot (data, aes(x=Catvar)) + geom_bar() + labs(title="", x="", y="")

ggplot(NHANES_cancer_project, aes(x=CancerType)) + geom_bar(fill="purple4") + labs(title="Cancer Type", x="Type of Cancer", y="Count")+ coord_flip()
```

```{r}
ggplot(NHANES_cancer_project, aes(x=CancerYN)) + geom_bar(fill="purple4") + labs(title="Cancer?", x="Y or N", y="Count")

ggplot(NHANES_cancer_project, aes(x=HowHealthy)) + geom_bar(fill="purple4") + labs(title="How Healthy?", x="Health", y="Count")

ggplot(NHANES_cancer_project, aes(x=AnnualHHInc)) + geom_bar(fill="purple4") + labs(title="Annual Household Income", x="", y="Count") + coord_flip()
```

```{r}
NHANES_cancer_project <- read.csv("C:/Users/acarb/Documents/NHANES download cancer/clean_data")
```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
install.packages("dunn.test")
library(dunn.test)
```

```{r}
#F-test: var.test(Contvar~catvar, data=, alternative="two.sided")
var.test(FamilyInc.Poverty~CancerYN, data=NHANES_cancer_project, alternative="two.sided")
#T-Test: t.test(Contvar~Catvar, var.equal=FALSE, data=, na.rm=TRUE)
t.test(FamilyInc.Poverty~CancerYN, var.equal=FALSE, data=NHANES_cancer_project, na.rm=TRUE)
#Not a significant difference, greater than .05
```
Not statistically significant

```{r}
var.test(AvgCigSmoked~CancerYN, data=NHANES_cancer_project, alternative="two.sided")
t.test(AvgCigSmoked~CancerYN, var.equal=FALSE, data=NHANES_cancer_project, na.rm=TRUE)
#Not a sig difference, more than .05
```
Not significantly significant

Kruskal Wallis:
```{r}
#ggplot(NHANESdata, aes(x=catvar, y=contvar)) + geom_boxplot(fill="green") + labs(x="BMI cat", y="calories consumed per day", title="Daily cateogirs by BMIcat")

ggplot(NHANES_cancer_project, aes(x=CancerType, y=AvgCigSmoked)) + geom_boxplot(fill="purple4") + labs(x="Cancer Type", y="Avg Cig Smoked/day", title="Cancer Type by Avg Cig Smoked/day")+ coord_flip()

#dunn.test(x=data$contvar, g=data$catvar, kw=TRUE)
dunn.test(x=NHANES_cancer_project$AvgCigSmoked, g=NHANES_cancer_project$CancerType, kw=TRUE)
```
People with Melanoma, Lung, Colon are most common with people who smoke the most.  

```{r}
ggplot(NHANES_cancer_project, aes(x=CancerType, y=FamilyInc.Poverty)) + geom_boxplot(fill="purple4") + labs(x="Cancer Type", y="Family Inc:Poverty", title="Cancer Type by FamilyInc:Poverty")+ coord_flip()

dunn.test(x=NHANES_cancer_project$FamilyInc.Poverty, g=NHANES_cancer_project$CancerType, kw=TRUE)
```
Thyroid and Bladder, most common 

```{r}
ggplot(NHANES_cancer_project, aes(x=AvgCigSmoked, y=FamilyInc.Poverty)) + geom_boxplot(fill="purple4") + labs(x="Avg Cig Smoked", y="Family Inc:Poverty", title="Avg Cig Smoked by FamilyInc:Poverty")+ coord_flip()
```

Chi Squared: 
```{r}
#a. Generate a two-way frequency table of BMIcat and HHFoodSec:
#table_name <- xtabs(~ CatVar1 + CatVar2, data=)
#print the table
#show the table in percentages: prop.table(table_name)*100
CancerYN_AnnualHHInc <- xtabs(~CancerYN + AnnualHHInc, NHANES_cancer_project)
CancerYN_AnnualHHInc
prop.table(CancerYN_AnnualHHInc)*100

#b. run the chi-square test:
#chisq.test(table_name)
chisq.test(CancerYN_AnnualHHInc)
```
Not a significant difference.  

```{r}
CancerYN_HowHealthy <- xtabs(~CancerYN + HowHealthy, NHANES_cancer_project)
CancerYN_HowHealthy
prop.table(CancerYN_HowHealthy)*100

chisq.test(CancerYN_HowHealthy)
```
Not a significant difference

```{r}
CancerType_AnnualHHInc <- xtabs(~CancerType + AnnualHHInc, NHANES_cancer_project)
CancerType_AnnualHHInc
prop.table(CancerType_AnnualHHInc)*100

chisq.test(CancerType_AnnualHHInc)
```

```{r}
CancerType_HowHealthy <- xtabs(~CancerType + HowHealthy, NHANES_cancer_project)
CancerType_HowHealthy
prop.table(CancerType_HowHealthy)*100

chisq.test(CancerType_HowHealthy)
```


Correlation
```{r}
library(dplyr)
library(ggplot2)
```

```{r}
#scatter plot code: #ggplot(aes(x=DepVar, y=IndVar), data=) + geom_point(color="") + geom_smooth(method=lm, color="") + labs(x="", y="") + theme_bw()
ggplot(aes(x=AvgCigSmoked, y=FamilyInc.Poverty), data=NHANES_cancer_project) + geom_point(color="purple4") + geom_smooth(method=lm, color="cornflowerblue") + labs(x="Avg Cig Smoked", y="Poverty Level") + theme_bw()

#correlation test code: cor.test(data$DepVar, data$IndVar, method="pearson")
cor.test(NHANES_cancer_project$AvgCigSmoked, y=FamilyInc.Poverty, method="pearson")
```
Higher the poverty level, the less an individual smoked

Correlation: 
```{r}
#model_name <- lm(DepVar~IndVar, data=)
#summary(model_name)

AvgCigSmoked_FamilyInc.Poverty_Fit <-lm(AvgCigSmoked~FamilyInc.Poverty, data=NHANES_cancer_project)
summary(AvgCigSmoked_FamilyInc.Poverty_Fit)
```
Statistically Significant 

```{r}
AvgCigSmoked_HowHealthy_fit <- lm(AvgCigSmoked~HowHealthy, data=NHANES_cancer_project)
summary(AvgCigSmoked_HowHealthy_fit)
```
Not statistically significant

```{r}
AvgCigSmoked_CancerType_fit <- lm(AvgCigSmoked~CancerType, data=NHANES_cancer_project)
summary(AvgCigSmoked_CancerType_fit)
```
Not statistically significant

```{r}
FamilyInc.Poverty_CancerYN_fit <- lm(FamilyInc.Poverty~CancerYN, data=NHANES_cancer_project)
summary(FamilyInc.Poverty_CancerYN_fit)

```
Not Statistically significant

Odds Ratio:
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
install.packages("epitools")
library(epitools)
install.packages("ggmosaic")
library(ggmosaic)
install.packages(scales)
library(scales)
```

```{r}
#1. Create the odds table:
#odds_table_name <- table(data$IndVar, data$DepVar)
HowHealthy_CancerType_OddsTable <- table(NHANES_cancer_project$HowHealthy, NHANES_cancer_project$CancerType)
#replaced 0s with 1s for oddsratio to work (malena did this)
HowHealthy_CancerType_OddsTable[HowHealthy_CancerType_OddsTable == 0] <- 1
#2. Run the oddsratio on the table you created:
#oddsratio(odds_table_name)
oddsratio(HowHealthy_CancerType_OddsTable)
#3. Graph the results in a mosaic table:
#ggplot(data=) + geom_mosaic(aes(x=product(IndVar), fill=DepVar), na.rm=TRUE, show.legend=FALSE) + labs(title="", x="", y="")
ggplot(data=NHANES_cancer_project) + geom_mosaic(aes(x=product(HowHealthy), fill=CancerType), na.rm=TRUE, show.legend=FALSE) + labs(title="", x="", y="") + coord_flip() + theme(axis.text.x = element_text(angle = 60, hjust = .5, vjust = .5, size = 6))
```

